A Constraint Solving Approach to Parikh Images of Regular Languages

### 2023-07-11 generalised.tex
more

Parikh images are also applicable for deciding more general classes of string
constraints~\cite{ostrich-plus}. Consider the substring constraint
\cref{const:s1-substring} of \cref{ex:string-constraints}: $ s_1 =
\Substring(s_2, i, n)$, that is $s_1$ is an $n$-length substring of $s_2$
starting at offset $i$ (in some variants , $n$ is $i + |n|$, i.e. the ending
index). That constraint belongs to an expressive fragment of string logic that
cannot be decided by most state-of-the-art string solvers.

In \cref{sec:intuition} we hand-wavily modelled \cref{const:s1-substring} (and
  the other constraints) using \emph{Parikh
  automata}~\cite{parikh-automata,expressiveness}. A Parikh automaton is a
  standard NFA extended such that transitions are additionally labelled both
  with offset vectors defining the increments of a finite number of counters.
  This means that Parikh automata recognise words over an extended
  alphabet~$\Alphabet \times D$, where $D \subseteq \Naturals^d$ is a finite set
  of increment vectors (notation as in~\cite{expressiveness}) and $\Alphabet$ is
  the alphabet of the original NFA.

@@

Applied to \cref{const:s1-substring}, the decision procedure in
\cite{ostrich-plus} will construct a pre-image of $s_1$ under the substring
operation, and check whether this pre-image is consistent with the constraint~$s_2 \in \Language(B)$, corresponding to the full product seen in
\cref{ex:string-constraints}. Because the substring operation depends on the
values of the integer variables~$i, n$, a Parikh automaton of dimension~$2$,
shown in \cref{fig:aut_a}, models the pre-image. Intuitively, this construction
accumulates any symbol (for the prefix of the substring operation), incrementing
$i$ to mark the start of the substring, followed by the automaton representing
the substring itself, modified to add a counter to increment $n$ at each
transition, followed by a state that performs no increments and only accepts any
suffix after the matched substring.

% AS: I am quite unsure about which intersection we are looking for here,
% please verify I'm not wrong!
% PR: Sounds right (did some small fixes)
Denoting the language described by \cref{fig:aut_a} as $\Language_{pre}$, we can
then replace $s_1 = \Substring(s_2, i, n)$ and $s_2 \in
\Language(B)$ (of \cref{const:s1-in-c-dd}) with an
equi-satisfiable formula that no longer contains any explicit substring
operation:
%
\begin{equation}
  \label{eq:parikh-constraint}
  p \in \Language_{pre} \wedge~
  \pi_\Alphabet(p) \in \Language(B)
  \wedge~~ \pi_D(p) =
  \begin{bmatrix}
    i \\ n
  \end{bmatrix}~~
  \wedge 0 \leq i \leq i + n \leq |p|
\end{equation}
%
To check the satisfiability of \cref{eq:parikh-constraint}, we need a
decision procedure that can process intersections of regular languages
(in this case, of $\Language_{pre}$ and $\Language(B)$, synchronising
on $\Alphabet$), while imposing the side
condition~$0 \leq n \leq m \leq |p|$ on the increment sum. In
\cite{ostrich-plus}, this decision procedure turned out to be main
bottleneck of the string solver, which was one of the motivations to
develop the lazy algorithm proposed in this paper.

>>>>

Parikh images are also applicable for deciding more general classes of string
constraints~\cite{ostrich-plus}. Consider the substring constraint
\cref{const:s1-substring} of \cref{ex:string-constraints}: $ s_1 =
\Substring(s_2, i, n)$, that is $s_1$ is an $n$-length substring of $s_2$
starting at offset $i$.
% (in some variants , $n$ is $i + |n|$, i.e. the ending index).
That constraint belongs to an expressive fragment of string logic that
cannot be decided by most state-of-the-art string solvers.

In \cref{sec:intuition} we modelled \cref{const:s1-substring} (and the
other constraints) using \emph{Parikh
  automata}~\cite{parikh-automata,expressiveness}. A Parikh automaton
is a finite-state automaton extended such that transitions are
additionally labelled with offset vectors defining the increments
of a finite number of counters.  This means that Parikh automata
recognise words over an extended alphabet~$\Alphabet \times D$, where
$D \subseteq \Naturals^d$ is a finite set of increment vectors
(notation as in~\cite{expressiveness}) and $\Alphabet$ is the alphabet
of the original automaton.

@@

Applied to \cref{const:s1-substring}, the decision procedure in
\cite{ostrich-plus} will construct a pre-image of $s_1$ under the substring
operation, and check whether this pre-image is consistent with the constraint~$s_2 \in \Language(B)$, corresponding to the full product seen in
\cref{ex:string-constraints}. Because $\Substring$ depends on the
values of the integer variables~$i, n$, a Parikh automaton,
shown in \cref{fig:aut_a}, models the pre-image. The Parikh automaton
has dimension~$3$, as it includes also the length register~$l_{A}$, besides
variables $i, n$.  Intuitively, this construction
accumulates any prefix symbol, incrementing
$i$ to mark the start of the substring; followed by the automaton representing
the substring itself, modified to add a counter to increment $n$ at each
transition; followed by a state that accepts any
suffix after the matched substring.

% AS: I am quite unsure about which intersection we are looking for here,
% please verify I'm not wrong!
% PR: Sounds right (did some small fixes)
Denoting the language described by \cref{fig:aut_a} as $\Language_{pre}$, we can
then replace $s_1 = \Substring(s_2, i, n)$ and $s_2 \in
\Language(B)$ (of \cref{const:s1-in-c-dd}) with an
equi-satisfiable formula that no longer contains any explicit substring
operation:
%
\begin{equation}
  \label{eq:parikh-constraint}
  p \in \Language_{pre} \wedge~
  \pi_\Alphabet(p) \in \Language(B)
  \wedge~~ \pi_D(p) =
  \begin{bmatrix}
    l_{A} \\ i \\ n
  \end{bmatrix}~~
  \wedge 0 \leq i \leq i + n \leq l_{A}
\end{equation}
%
To check the satisfiability of \cref{eq:parikh-constraint}, we need a
decision procedure that can process intersections of regular languages
(in this case, of $\Language_{pre}$ and $\Language(B)$, synchronising
on $\Alphabet$), while imposing the side
condition~$0 \leq n \leq m \leq |p|$ on the increment sum. In
\cite{ostrich-plus}, this decision procedure turned out to be main
bottleneck of the string solver, which was one of the motivations to
develop the lazy algorithm proposed in this paper.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
